\MakeUppercase{Linear versus centred chromatic numbers}\thanks{This research was partly funded by NSERC.}

### 2022-05-11 lin-vs-cen.tex
Pfffft

\begin{lem}\label{packing_lemma}
  Let $r\ge 2$ be an even integer and let $Q\subseteq \VE(\interior_r(G_k))$ be such that $\mu_1\not\in \VE(G_r(\mu_2))$ for each distinct $\mu_1,\mu_2\in Q$.  Then, for any $\mu\in \VE(G_k)$,
  $|\{\mu_1\in Q: \mu\in \VE(G_{2r}(\mu))\}| \le 25$.
\end{lem}

\begin{proof}
  The packing condition that $\mu_1\not\in \VE(G_r(\mu_2))$ for each distinct $\mu_1,\mu_2\in Q$ implies that $\VE(G_{r/2-1}(\mu_1))\cap \VE(G_{r/2-1}(\mu_2))=\emptyset$ for each $\mu_1,\mu_2\in Q$.  Let $Q':=\{\mu_1\in Q: \mu\in \VE(G_{2r}(\mu_1))\}$ so that our task is to show that $|Q'|\le 25$.  For any $\mu_1\in Q'$,
  $\mu_1\in\VE(G_{2r+1}(\mu))$ and consequently, $\VE(G_{r/2-1}(\mu_1))\subseteq \VE(G_{5r/2}(\mu))$.
  Since this is true for all $\mu_1\in Q$ we have
  \[  \vol{5r/2} \ge |\VE(G_{5r/2}(\mu))| \ge \sum_{\mu_1\in Q'} |\VE(G_{r/2}(\mu_1))|
  = |Q'|\cdot \vol{r/2} \enspace .
  \]
%   /* \begin{lstlisting}[frame=single,language=python]
%     # Sage Code
%   g(r)=2*(r+1)*(2*r+1)  # |B_r(mu)|
%   show(g)
%   show(g.factor())
%   show(g.full_simplify())
%   f(r)=2*(r+1)*(2*r+1 + 2*r) + (2*r+1)^2  # |VE(G[B_r(mu)])
%   show(f)
%   show(f.factor())
%   vol = f
%   ratio(r) = vol(5*r/2) / vol(r/2)  # ratio that appears in packing lemma
%   show(plot(ratio, r, 1, 10))
%   show(limit(ratio(r), r=oo))
%   \end{lstlisting}
  which implies that $|Q'|\le \vol{5r/2}/\vol{r/2}$.  For $r > 2$, this bound is increasing in $r$ and is less than $25$ for any $r\ge 2$.
\end{proof}

@@

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 1$ there exists an integer $d\in O(r^4\log r)$ such that the following is true, for every integer $k\ge 1$.\newline
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)|\ge d|A|$.
  Then there exists $S\subseteq V(\interior_r(G))$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$ for each $v\in S$.
  \end{compactenum}
\end{lem}

@@

  We construct $Q$ in two rounds.  In the first round, we start with an initially empty set $Q_1$ and repeat the following for each $\alpha\in\varphi(V(G))$:
  If there exists distinct $\mu_1,\mu_2\in\VE(G_k)$ such that
  \begin{compactenum}[(a)]
    \item $\phi(\mu_1)=\phi(\mu_2)=\alpha$; and\label{hits_both_q}
    \item for each $b\in\{1,2\}$, $\mu_b\not\in \VE(G_r(\mu_{3-b})) \cup \bigcup_{\mu\in Q_1} \VE(G_r(\mu))$,\label{spread_out_q}
  \end{compactenum}
  then we include $\mu_1$ and $\mu_2$ in $Q_1$ and declare the first round an \defin{$\alpha$-success}.  Otherwise, we declare the first round an \defin{$\alpha$-failure}.

  Let $\overline{X}\subseteq\varphi(V(G))$ be the set of colours $\alpha$ for which the first round was an $\alpha$-success. At the end of this process, the set $Q_1$ certainly satisfies (\ref{spread_out}) since (\ref{spread_out_q}) implies that $\VE(G_r(\mu))\cap Q_1=\{\mu\}$ for each $\mu\in Q_1$.  However, $Q_1$ only satisfies (\ref{hits_both}) for the colours in $\overline{X}$.  We now use the second round to create a set $Q_2$ to fix this.
  % Our strategy is to use \cref{d_hall} to find two matchings between the colours in $X$ and the boxes in $\{\tilde{B}_r(\mu):\mu\in Q_1\}$ that contain those colours.

  Define a bipartite graph $H$ with parts $L(H):=X:=\varphi(V(G))\setminus \overline{X}$ and $R(H):=Y:=Q_1$.  Let $\pi:=\mu_1,\ldots,\mu_{|Y|}$ be a random permutation of $Y$.  We include an edge $\alpha\mu_i$ in $H$ if and only if
  \[
    \VE(G_r(\mu_i))\setminus\left(\bigcup_{j=1}^{i-1}\VE(G_{2r}(\mu_j))\right)
  \]
  contains some object $\nu$ of colour $\phi(\nu)=\alpha$.  Note the asymmetry here: In the first round, each element $\mu_j$ that takes part in $Q_1$ prevents elements in $\VE(G_r(\mu))$ from also being part of $Q_1$.  In the second round, $\mu_j$ defines a larger box $\VE(G_{2r}(\mu_j))$ that must be avoided when choosing a second element from $\VE(G_r(\mu_i))$, for each $i>j$.

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{r}$ for each $\alpha\in X$.
  For each $\alpha\in X$, consider the set $\phi^{-1}_{\mathcal{P}}(\alpha)\subseteq \VE(G_k)$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\phi^{-1}_{\mathcal{P}}(\alpha)| \ge d$.  Furthermore all but at most $\vol{r}$ elements of $\phi^{-1}_{\mathcal{P}}(\alpha)$ are contained in $\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))$ since otherwise the first round would have been an $\alpha$-success.  Let $\Phi_\alpha:=\phi^{-1}_{\mathcal{P}}(\alpha)\cap \left(\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))\right)$, so $|\Phi_\alpha|\ge d-\vol{2r+2}$.  In words, $\Phi_\alpha$ consists of all objects of colour $\alpha$ that could not be chosen in the first round because they are too close to an element of $Y$.

@@

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\ceil{|\Phi_\alpha|/\vol{10r}},1/25)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{r})$.  (We make no attempt to optimize constants here.)  Suppose that, for some $\mu\in Y$, $B_r(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$. Then, by \cref{packing_lemma}, the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{2r}(\xi))\}
  \]
  has size at most $25$. Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  This happens with probability exactly $1/|Y_\nu|\ge 1/25$. This is already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$, but in order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.\footnote{There are many ways to obtain a concentration result here, we use a simple method with the goal of being self-contained.  For a tighter result, use the method of bounded differences \cite{mcdiarmid:on}.}  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_r(\mu))\cap J$ contains at least one element $\nu$ and then set $J:=J-\VE(G_{10r}(\mu))$.\footnote{The constant $10$ is overkill here and is only used for simplicity; $8r+O(1)$ would be sufficient.}  This process continues until $\VE(G_r(\mu))\cap J=\emptyset$ for each $\mu\in Y$. Since $|\VE(G_{10r}(\mu))|\le \vol{10r}$, each step in this process eliminates at most $\vol{10r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{10r}}$.

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the  permutation $\pi$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/25)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{r}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{r}\right) \notag \\
    & \le \Pr(\binomial(t,1/25)\le 2\vol{r}) \notag \\
    & = \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{1}{25}\right)^x\left(\frac{24}{25}\right)^{t-x} \notag \\
    & < \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{24}{25}\right)^{t} \notag \\
    & < (2t)^{2\vol{r}}\cdot\left(\frac{24}{25}\right)^{t} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & = \exp\big({2\vol{r}}\ln(2t)-t\ln(25/24)\big) \notag \\
    & \le \exp\left({2\vol{r}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|\ln(25/24)}{\vol{10r}}\right)
      & \text{(since $|\Phi_\alpha|/\vol{10r} \le t \le |\Phi_\alpha|$)} \notag \\
    & \le \exp\left({2\vol{r}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|}{25\vol{10r}}\right)
      & \text{(since $\ln(25/24) > 1/25$).}
       \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

  We are now ready for an application of \cref{weighted_lovasz}.  For each $\alpha\in X$, let $E_\alpha$ denote the event ``$\deg_H(\alpha)\le 2\vol{r}$'' and let $\mathcal{E}:=\{E_\alpha:\alpha\in X\}$. For each $\alpha\in X$, define
  \[
     \Gamma_\alpha := \left\{E_\beta \in \mathcal{E}: \Phi_\beta\cap \left(\bigcup_{\nu\in \Phi_\alpha} \VE(G_{10r}(\nu))\right)\neq\emptyset\right\}
  \]
  and observe that $|\Gamma_\alpha|\le \vol{10r}\cdot|\Phi_\alpha|$.
  The event $E_\alpha$ is mutually independent of $\mathcal{E}\setminus\Gamma_\alpha$ since, for any $E_\beta\in\mathcal{E}\setminus\Gamma_\alpha$ the sets $Y_\alpha:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\alpha\neq\emptyset\}$ and $Y_\beta:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\beta\neq\emptyset\}$ are disjoint and $\deg_H(\alpha)$ and $\deg_H(\beta)$ are each determined by the relative orders of $Y_\alpha$ and $Y_\beta$ in $\pi$, respectively.

  For each $\alpha\in X$, let $w(E_\alpha):=1/\tau$, for some $\tau>0$ to be determined shortly.  Then
  \begin{align}
    w(E_\alpha)\prod_{E_\beta\in\Gamma_\alpha} (1-w(E_\beta))
    & = \tfrac{1}{\tau}\left(1-\tfrac{1}{\tau}\right)^{|\Gamma_\alpha|-1} \notag \\
    & \ge \exp\left(-\ln\tau - |\Gamma_\alpha|/(\tau-1)\right) & \text{(Since $(1-1/\tau)^{\tau-1} > 1/e$)} \notag \\
    & \ge \exp\left(-\ln\tau - \vol{10r}\cdot|\Phi_\alpha|/(\tau-1)\right) \enspace . \label{weight_product}
  \end{align}

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {50\vol{10r}}+1$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(\Gamma_\alpha))
  \]
  provided that
  \[   |\Phi_\alpha| \ge {50\vol{10r}}\big(\ln({50\vol{10r}}+1) + {2\vol{r}}\ln(2|\Phi_\alpha|)\big) \enspace .
  \]
  The right hand side of this last equation is of the form $O(r^2+r^4\log|\Phi_\alpha|)$.  Since $|\Phi_\alpha|\ge d$, this can therefore be satisfied for some $d\in O(r^4\log r)$.

  % Taking the ratio of \cref{ratioo} and $\Pr(E_\alpha)$ we obtain
  % \begin{align*}
  %   \frac{e^{-C|\Phi_\alpha|/2r^2}-e^{-C|(\Gamma_\alpha|+1)/50r^2}}{\Pr(E_\alpha)}
  %   & \ge \frac{e^{-C|\Phi_\alpha|/2r^2}-e^{-C|(\Gamma_\alpha|+1)/50r^2}}{e^{-C|\Phi_\alpha|/r^2}} \\
  %   & \ge e^{C|\Phi_\alpha|/2r^2} - \frac{e^{-C|(\Gamma_\alpha|+1)/50r^2}}{e^{-C|\Phi_\alpha|/r^2}} \\
  % \end{align*}
  %
  % \todo[inline]{Hmmmmm. Not working.  Try something like $w(E_\alpha)=1-e^{-1/50}$}
  %
  % Finally, by \cref{packing_lemma} $|\Gamma_\alpha|\le 25|\Phi_\alpha|$.
  %
  % \begin{align*}
  %     \Pr(E_\alpha) \le \exp(-|\Phi_\alpha|\cdot C/r^2) \enspace .
  % \end{align*}
  %
  % For each $\alpha\in X$, define $w(E_\alpha)=C/2r^2$ and observe that
  % \begin{align*}
  %   p(1-p)^{|\Gamma_\alpha|}
  %   & \ge  p(1-p)^{25|\Phi_\alpha|} = p\exp(|\Phi_\alpha|\cdot 25\log(1-c/2r^2))
  % \end{align*}
  % \todo[inline]{Sort out these calculations}

@@

\begin{proof}[Proof of \cref{pseudogrid_lower_bound}]
  Suppose for the sake of contradiction that $G$ has a linear colouring $\varphi$ using fewer than $\epsilon k$ colours for arbitrarily small $\epsilon >0$.  Then, by \cref{only_frequent}, $G$ contains a $k'\times k'$ pseudogrid $G'$ with $k'\ge k/2$ having a grid-decomposition $\mathcal{P}':=\{P_\mu':\mu\in \VE(G_{k'})\}$ with the property that, for any $A\subseteq\varphi(V(G'))$, $|\varphi^{-1}_{\mathcal{P'}}(A)\cap\interior_r(G_{k'})|\ge d|A|$, for any $d\le 1/2\epsilon-2$.  In particular, for some $\epsilon \in \Omega(1/r^4)$, \cref{doubled_colour_set} implies that $G'$ contains a set $S$ of $2|\varphi(V(G')|$ vertices containing two vertices of each colour in $\varphi(V(G'))$ such that $|B_r(\mu)\cap S|\le 2$ for each $\mu\in S$.  By \cref{pick_up_everything}, $G'$ contains a path $P$ that contains every vertex in $S$.  Since $P$ is contained in $G'$ $\varphi(V(P))=\varphi(V(G'))$, so $P$ is has no center under $\varphi$, contradicting the assumption that $\varphi$ is a linear colouring of $G$.
\end{proof}

>>>>

\begin{lem}\label{packing_lemma}
  Let $r\ge 10$ be an even integer and let $Q\subseteq \VE(G_k)$ be such that $\mu_1\not\in \VE(G_{2r+1}(\mu_2))$ for each distinct $\mu_1,\mu_2\in Q$.  Then, for any $\mu\in \VE(G_k)$,
  $|\{\mu_1\in Q: \mu\in \VE(G_{3r+1}(\mu_1))\}| \le 16$.
\end{lem}

\begin{proof}
  To avoid the need to discuss boundary conditions that have no effect on the result, it is simpler to consider point that are sufficiently far from the boundary of $G_k$. More precisely, we may assume that $Q\cup\{\mu\}\subseteq\interior_{4r+2}(G_{k+8r+4})$.

  The packing condition that $\mu_1\not\in \VE(G_{2r+1}(\mu_2))$ for each distinct $\mu_1,\mu_2\in Q$ implies that $\VE(G_{r}(\mu_1))\cap \VE(G_{r}(\mu_2))=\emptyset$.  Let $Q':=\{\mu_1\in Q: \mu\in \VE(G_{3r+1}(\mu_1))\}$ so that our task is to show that $|Q'|\le 16$.  For any $\mu_1\in Q'$,
  $\mu_1\in\VE(G_{3r+2}(\mu))$ and consequently, $\VE(G_{r}(\mu_1))\subseteq \VE(G_{4r+2}(\mu))$.  For each $\mu_1\in Q'$, $|\VE(G_r(\mu_1))|= 12r^2+8r+1$.  Therefore,
  \[  |Q'|\le \frac{\vol{4r+2}}{12r^2+8r+1} < 17 \enspace ,
  \]
  for all $r\ge 10$.
\end{proof}

@@

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 10$ there exists an integer $d\in O(r^4\log r)$ such that the following is true, for every integer $k\ge 1$.\newline
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)|\ge d|A|$.
  Then there exists $S\subseteq V(\interior_r(G))$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$ for each $v\in S$.
  \end{compactenum}
\end{lem}

@@

  We construct $Q$ in two rounds.  In the first round, we start with an initially empty set $Q_1$ and repeat the following for each $\alpha\in\varphi(V(G))$:
  If there exists distinct $\mu_1,\mu_2\in\VE(G_k)$ such that
  \begin{compactenum}[(a)]
    \item $\phi(\mu_1)=\phi(\mu_2)=\alpha$; and\label{hits_both_q}
    \item for each $b\in\{1,2\}$, $\mu_b\not\in \VE(G_{2r+1}(\mu_{3-b})) \cup \bigcup_{\mu\in Q_1} \VE(G_{2r+1}(\mu))$,\label{spread_out_q}
  \end{compactenum}
  then we include $\mu_1$ and $\mu_2$ in $Q_1$ and declare the first round an \defin{$\alpha$-success}.  Otherwise, we declare the first round an \defin{$\alpha$-failure}.

  Let $\overline{X}\subseteq\varphi(V(G))$ be the set of colours $\alpha$ for which the first round was an $\alpha$-success. At the end of this process, the set $Q_1$ certainly satisfies (\ref{spread_out}).  In fact, it implies an even stronger property:  $|B_r(\nu)\cap Q_1|\le 1$ for any $\nu\in\VE(G_k)$. However, $Q_1$ only satisfies (\ref{hits_both}) for the colours in $\overline{X}$.  We now use the second round to create a set $Q_2$ to handle the colours in $X:=\varphi(V(G))\setminus \overline{X}$.
  % Our strategy is to use \cref{d_hall} to find two matchings between the colours in $X$ and the boxes in $\{\tilde{B}_r(\mu):\mu\in Q_1\}$ that contain those colours.

  Define a bipartite graph $H$ with parts $L(H):=X$ and $R(H):=Y:=Q_1$.  Let $\pi:=\mu_1,\ldots,\mu_{|Y|}$ be a random permutation of $Y$.\footnote{Here random permutation means a permutation chosen uniformly over all $|Y|!$ permutations of $Y$.}  We include an edge $\alpha\mu_i$ in $H$ if and only if
  \[
    \VE(G_r(\mu_i))\setminus\left(\bigcup_{j=1}^{i-1}\VE(G_{3r+1}(\mu_j))\right)
  \]
  contains some object $\nu$ of colour $\phi(\nu)=\alpha$.  Note the asymmetry here: In the first round, each element $\mu_j$ that takes part in $Q_1$ prevents elements in $\VE(G_r(\mu))$ from also being part of $Q_1$.  In the second round, $\mu_j$ defines a larger box $\VE(G_{3r+1}(\mu_j))$ that must be avoided when choosing a second element from $\VE(G_{2r+1}(\mu_i))$, for each $i>j$.

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{2r+1}$ for each $\alpha\in X$.
  For each $\alpha\in X$, consider the set $\phi^{-1}_{\mathcal{P}}(\alpha)\subseteq \VE(G_k)$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\phi^{-1}(\alpha)| \ge d$.  Furthermore all but at most $\vol{2r+1}$ elements of $\phi^{-1}(\alpha)$ are contained in $\bigcup_{\mu\in Y}\VE(G_{2r+1}(\mu))$ since otherwise the first round would have been an $\alpha$-success.  Let $\Phi_\alpha:=\phi^{-1}(\alpha)\cap \left(\bigcup_{\mu\in Y}\VE(G_{2r+1}(\mu))\right)$, so $|\Phi_\alpha|\ge d-\vol{2r+2}$.  In words, $\Phi_\alpha$ consists of all objects of colour $\alpha$ that were eliminated from consideration because they are too close to an element of $Y$.

@@

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\ceil{|\Phi_\alpha|/\vol{10r}},1/16)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{r})$.  (We make no attempt to optimize constants here.)  Suppose that, for some $\alpha\in X$ and $\mu\in Y$, $B_{2r+1}(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$ and consider the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{3r+1}(\xi))\} \enspace .
  \]
  Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  By \cref{packing_lemma}, $|Y_\nu|\le 16$, so the probability that the edge $\alpha\mu$ appears in $H$ is at least $1/|Y_\nu|\ge 1/16$. This is already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$, but in order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.\footnote{There are many ways to obtain a concentration result here, we use a simple method with the goal of being self-contained.  For a tighter result, use the method of bounded differences \cite{mcdiarmid:on}.}  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_r(\mu))\cap J$ contains at least one element $\nu$ and then set $J:=J\setminus \VE(G_{7r}(\nu))$.\footnote{The constant $7$ is overkill here and is only used for simplicity; $6r+O(1)$ would be sufficient.}  This process continues until $\VE(G_r(\mu))\cap J=\emptyset$ for each $\mu\in Y$. Since $|\VE(G_{7r}(\mu))|\le \vol{7r}$, each step in this process eliminates at most $\vol{7r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{7r}}$.

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_t\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are pairwise disjoint.  Indeed, $Y_{\nu_i}\subseteq B_{3r+2}(\nu_i)$, $Y_{\nu_j}\subseteq B_{3r+2}(\nu_j)$ and the fact that $\nu_i\not\in B_{7r}(\nu_j)$ implies that $B_{3r+2}(\nu_i)\cap B_{3r+2}(\nu_j)=\emptyset$.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_t$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of the subset $Y_{\nu_i}$ in the permutation $\pi$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/16$.  Therefore, $\deg_H(\alpha)\ge \sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/16)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{r}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{r}\right) \notag \\
    & \le \Pr(\binomial(t,1/16)\le 2\vol{r}) \notag \\
    & = \sum_{x=0}^{2\vol{2r+1}} \binom{t}{x}\left(\frac{1}{16}\right)^x\left(\frac{15}{16}\right)^{t-x} \notag \\
    & < \sum_{x=0}^{2\vol{2r+1}} \binom{t}{x}\left(\frac{15}{16}\right)^{t} \notag \\
    & < (2t)^{2\vol{2r+1}}\cdot\left(\frac{15}{16}\right)^{t} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & = \exp\big({2\vol{2r+1}}\ln(2t)-t\ln(16/15)\big) \notag \\
    & \le \exp\left({2\vol{2r+1}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|\ln(16/15)}{\vol{10r}}\right)
      & \text{(since $|\Phi_\alpha|/\vol{10r} \le t \le |\Phi_\alpha|$)} \notag \\
    & \le \exp\left({2\vol{2r+1}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|}{16\vol{10r}}\right)
      & \text{(since $\ln(16/15) > 1/16$).}
       \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

  We are now ready for an application of \cref{weighted_lovasz}.  For each $\alpha\in X$, let $E_\alpha$ denote the event ``$\deg_H(\alpha)\le 2\vol{2r+1}$'' and let $\mathcal{E}:=\{E_\alpha:\alpha\in X\}$. For each $\alpha\in X$, define
  \[
     \Gamma_\alpha := \left\{E_\beta \in \mathcal{E}: \Phi_\beta\cap \left(\bigcup_{\nu\in \Phi_\alpha} \VE(G_{7r}(\nu))\right)\neq\emptyset\right\}
  \]
  and observe that $|\Gamma_\alpha|\le \vol{7r}\cdot|\Phi_\alpha|$.
  The event $E_\alpha$ is mutually independent of $\mathcal{E}\setminus\Gamma_\alpha$ since, for any $E_\beta\in\mathcal{E}\setminus\Gamma_\alpha$ the sets $Y_\alpha:=\{\mu\in Y:B_{3r+1}(\mu)\cap \Phi_\alpha\neq\emptyset\}$ and $Y_\beta:=\{\mu\in Y:B_{3r+1}(\mu)\cap \Phi_\beta\neq\emptyset\}$ are disjoint and $\deg_H(\alpha)$ and $\deg_H(\beta)$ are each entirely determined by the relative orders of $Y_\alpha$ and $Y_\beta$ in $\pi$, respectively.

  For each $\alpha\in X$, let $w(E_\alpha):=1/\tau$, for some $\tau>1$ to be defined shortly.  Then
  \begin{align}
    w(E_\alpha)\prod_{E_\beta\in\Gamma_\alpha} (1-w(E_\beta))
    & = \tfrac{1}{\tau}\left(1-\tfrac{1}{\tau}\right)^{|\Gamma_\alpha|-1} \notag \\
    & \ge \exp\left(-\ln\tau - |\Gamma_\alpha|/(\tau-1)\right) & \text{(Since $(1-1/\tau)^{\tau-1} > 1/e$)} \notag \\
    & \ge \exp\left(-\ln\tau - \vol{7r}\cdot|\Phi_\alpha|/(\tau-1)\right) \enspace . \label{weight_product}
  \end{align}

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {32\vol{7r}}+1$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(\Gamma_\alpha))
  \]
  provided that
  \[   |\Phi_\alpha| \ge {32\vol{7r}}\big(\ln({32\vol{7r}}+1) + {2\vol{2r+1}}\ln(2|\Phi_\alpha|)\big) \enspace .
  \]
  The right hand side of this last equation is of the form $O(r^2+r^4\log|\Phi_\alpha|)$.  Since $|\Phi_\alpha|\ge d$, this can therefore be satisfied for some $d\in O(r^4\log r)$.

@@

\begin{proof}[Proof of \cref{pseudogrid_lower_bound}]
  Suppose for the sake of contradiction that $G$ has a linear colouring $\varphi$ using fewer than $\epsilon k$ colours for arbitrarily small $\epsilon >0$.  Then, by \cref{only_frequent}, $G$ contains a $k'\times k'$ pseudogrid $G'$ with $k'\ge k/2$ having a grid-decomposition $\mathcal{P}':=\{P_\mu':\mu\in \VE(G_{k'})\}$ with the property that, for any $A\subseteq\varphi(V(G'))$, $|\varphi^{-1}_{\mathcal{P'}}(A)\cap\interior_r(G_{k'})|\ge d|A|$, for any $d\le 1/2\epsilon-2$.  In particular, for some $\epsilon \in \Omega(1/r^4\log r)$, \cref{doubled_colour_set} implies that $G'$ contains a set $S$ of $2|\varphi(V(G')|$ vertices containing two vertices of each colour in $\varphi(V(G'))$ such that $|B_r(\mu)\cap S|\le 2$ for each $\mu\in S$.  By \cref{pick_up_everything}, $G'$ contains a path $P$ that contains every vertex in $S$.  Since $P$ is contained in $G'$, $\varphi(V(P))=\varphi(V(G'))$, so $P$ is has no center under $\varphi$, contradicting the assumption that $\varphi$ is a linear colouring of $G$.
\end{proof}
